### 2014-10-21 doc/thesis/system.tex
Improving the text about subtyping rules

In this chapter we present the formalization of Typed Lua.
Besides its practical contribution, Typed Lua also has some interesting
contributions on the field of optional type systems for scripting
languages.
Namely, it is a simple type system that covers several Lua idioms
and features, such as the refinement of tables, multiple return values,
and optional parameters.
We use typing rules to explain how we implemented the features of Typed Lua.

@@

Our type system uses subtyping \citep{cardelli1984smi,abadi1996to} to order
types and consistent-subtyping \citep{siek2007objects,siek2013mutable}
to allow the interaction between statically and dynamically typed code.
We will explain the subtyping and consistent-subtyping rules along this
section while we introduce each type from our type system.
We will focus the discussion on the definition of subtyping because,
as we mentioned in Chapter \ref{chap:review}, we achieve the definition of
consistent-subtyping combining consistency and subtyping.
The differences between subtyping and consistent-subtyping are the way
they handle the dynamic type, and the fact that subtyping is transitive,
but consistent-subtyping is not.

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-level types}}\\
t ::= & \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; p \rightarrow r & \textit{function types}\\
& | \; \{v_{i}, ..., v_{n}\}_{closed|open|unique} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; \pi_{i} & \textit{projection types}\\
l ::= & \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
b ::= & \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
v ::= & \;\; k:t \; | \; \Const \; k:t & \textit{field types}\\ 
k ::= & \;\; l \; | \; b \; | \; \Any & \textit{key types}\\
\multicolumn{3}{c}{}\\
\multicolumn{3}{c}{\textbf{Second-level types}}\\
s ::= & \;\; t &\\
& | \; t* & \textit{variadic types}\\
& | \; t \times s & \textit{tuple types}\\
p ::= & \;\; \Void \; | \; s & \textit{parameters list type}\\
r ::= & \;\; \Void \; | \; s \; | \; s \sqcup r & \textit{return types}\\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

Literal types represent the type of literal constants.
They can be the boolean values $\False$ and $\True$,
any integer value, any floating point value, and any string value.
We shall see that literal types are important to type tables as records.

Our type system also includes the following base types: $\Boolean$,
$\Integer$, $\Number$, and $\String$.
The base types $\Boolean$ and $\String$ represent the values that
during run-time Lua tags as \texttt{boolean} and \texttt{string},
respectively;
while the base types $\Integer$ and $\Number$ represent the values
that during run-time Lua tags as \texttt{number}.
We introduced two different base types to represent Lua numbers
because Lua introduced integers recently.

The subtyping rules for literal types include a reflexive rule
and the rules for defining that literal types are subtypes of
their respective base types;
the subtyping rules for base types include a reflexive rule
and another rule for defining that $\Integer$ is subtype of
$\Number$.
We defined these subtyping rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-LITERAL}\\
\senv \vdash l \subtype l
\end{array}
\;
\begin{array}{c}
\mylabel{S-FALSE}\\
\senv \vdash \False \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-TRUE}\\
\senv \vdash \True \subtype \Boolean
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-INT1}\\
\senv \vdash {\it int} \subtype \Integer
\end{array}
\;
\begin{array}{c}
\mylabel{S-INT2}\\
\senv \vdash {\it int} \subtype \Number
\end{array}
\;
\begin{array}{c}
\mylabel{S-FLOAT}\\
\senv \vdash {\it float} \subtype \Number
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-STRING}\\
\senv \vdash {\it string} \subtype \String
\end{array}
\begin{array}{c}
\mylabel{S-BASE}\\
\senv \vdash b \subtype b
\end{array}
\;
\begin{array}{c}
\mylabel{S-NUMBER}\\
\senv \vdash \Integer \subtype \Number
\end{array}
\end{array}
\]

@@

Since subtyping is reflexive and transitive,
we could have omitted the reflexive rules \textsc{S-LITERAL} and
\textsc{S-BASE}, as well as the rule \textsc{S-INT2}.
More precisely, we could have used subsumption rules that define
the reflexiveness and transitiveness of subtyping.
In the case of \textsc{S-INT2}, we would not need this rule because
we could use the transitive subsumption rule to derive that
\[
\dfrac{\senv \vdash 1 \subtype \Integer \;\;\;
       \senv \vdash \Integer \subtype \Number}
      {\senv \vdash 1 \subtype \Number}
\]

@@

The type $\Value$ is the top type, so any first-level type is a
subtype of $\Value$.
We shall see that this type along with variadic types help to
preserve the semantics of Lua to drop extra values on multiple
assignments, and function calls.
Following we show its subtyping rule:
\[
\begin{array}{c}
\mylabel{S-VALUE}\\
\senv \vdash t \subtype \Value
\end{array}
\]

The dynamic type $\Any$ is the type that allows programmers to mix
static and dynamic typing.
The type $\Any$ is neither the bottom nor the top type, but a
separated typed that is a subtype only of itself:
\[
\begin{array}{c}
\mylabel{S-ANY}\\
\senv \vdash \Any \subtype \Any
\end{array}
\]

@@

The rules that handle the dynamic type are the only rules that
differ between subtyping and consistent-subtyping.
If we set the type $\Any$ as both bottom and top types of our
subtyping relation, then all programs would type check without errors.
This would happen due to the transitivity of subtyping, that is,
we would be able to down-cast any type to $\Any$ and then up-cast
$\Any$ to any type.
While the rule \textsc{S-ANY} is neutral and makes transitivity
not be a problem, the rules \textsc{C-ANY1} and \textsc{C-ANY2}
are the rules that allow the dynamic type to interact with other
first-level types, and thus allow dynamically typed code coexist
with statically typed code.
Consistent-subtyping cannot be transitive because of these rules,
as transitivity combined with them would make all programs compile
without type errors.

@@

Union types are disjoint, that is, the types in the union are not
consistent-subtype of any other type in the same union.
For instance, the union type $\Boolean \;\cup\; \Any$ results the
type $\Any$, because $\Boolean$ is consistent-subtype of $\Any$.
The union type $\Number \;\cup\; \Nil \;\cup\; 1$ results the union
type $\Number \;\cup\; \Nil$ because $1 \subtype \Number$.

\emph{/* Explain how Typed Lua refine union types */}

To introduce function types we first need to introduce
second-level types, as they appear on both sides of function types.
Basically, second-level types are tuples of first-level types
that can optionally end with a variadic type.
Our type system needs second-level types because tuples are
not first-class values in Lua, though they may appear on
argument passing, multiple returns, and multiple assignments.
A variadic type $t*$ is a generator for a sequence of values of
type $t \cup \Nil$, and it represents the type of the vararg
expression.
Next we show their subtyping rules:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VARARG}\\
\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
      {\senv \vdash t_{1}* \subtype t_{2}*}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TUPLE1}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2} \;\;\;
       \senv \vdash s_{1} \subtype s_{2}}
      {\senv \vdash t_{1} \times s_{1} \subtype t_{2} \times s_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TUPLE2}\\
\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \;\;\;
       \senv \vdash t_{1}* \subtype s_{2}}
      {\senv \vdash t_{1}* \subtype t_{2} \times s_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TUPLE3}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil \;\;\;
       \senv \vdash s_{1} \subtype t_{2}*}
      {\senv \vdash t_{1} \times s_{1} \subtype t_{2}*}
\end{array}
\end{array}
\]

The subtyping rules for variadic and tuple types show that
these types are covariant.
The subtyping rule for variadic types is straightforward:
it shows that a variadic type $t_{1}*$ is subtype of another
variadic type $t_{2}*$ if $t_{1} \cup \Nil$ is subtype of
$t_{2} \cup \Nil$.
The first rule for tuple types shows that a tuple type
$t_{1} \times s_{1}$ is subtype of another tuple type
$t_{2} \times s_{2}$ if $t_{1}$ is subtype of $t_{2}$
and $s_{1}$ is subtype of $s_{2}$.
As we mentioned, a tuple can optionally end with a variadic
type, and the second and third rules for tuple types handle
this case.
The second rule shows that a variadic type is subtype of
a tuple whenever the type of the value that it generates is subtype
of all the elements of the tuple.
The third rule shows that a tuple is subtype of a variadic
type whenever all the elements of the tuple are subtype of the
type of the value that it generates.
We shall see that these subtyping relations along with the
types $\Value$ and $\Nil$ allow our type system to reflect
the semantics of Lua on discarding extra parameters and
replacing missing parameters.

In our type system function types have the form $p \rightarrow r$,
where $p$ is the type of the parameters list and $r$ is the
return type.
The type of the parameters list can be $\Void$ or a tuple type,
while the return type can be $\Void$, a tuple type, or a union of tuples.
The type $\Void$ means an empty tuple.
Our type system includes union of tuples on the return type because
in Lua it is common to overload the return type to denote error,
as we mentioned in Section \ref{sec:statistics}.

@@

Our type system includes table types to represent Lua tables.
The syntactical form of table types is $\{ v_{i}, ..., v_{n} \}_{tag}$.
When $n = 0$, then $i = 0$ and it represents the type of an empty table.
When $n > 0$, then $i$ varies from 1 to $n$, and each $v_{i}$
represents the type of a table field.
Table fields represent maps from keys to values, and we restricted
the type of keys to literal types, base types, and the dynamic type;
but we can use any first-level type to represent the type of values.
We made this restriction to the type of keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The dynamic type is an option when we need a more loosen table type.

@@

The following subtyping rules show that \emph{closed} table types
have width subtyping, and depth subtyping on $\Const$ fields:

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash v_{j} \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{closed} \subtype \{v_{1}, ..., v_{n}\}_{closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2} \;\;\;
       \senv \vdash t_{2} \subtype t_{1}}
      {\senv \vdash k_{1}:t_{1} \subtype k_{2}:t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \Const \; k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\end{array}
\]

First, we had only \emph{closed} table types.
However, programmers often define tables that include optional fields,
and the subtyping rules we introduced do not let programmers to use
the table constructor to initialize table types that include optional
fields.
For instance, consider the table type we introduced in
Chapter 3:
\begin{align*}
\{ & ``firstname":\String, \\
   & ``middlename":\String \cup \Nil,\\
   & ``lastname":\String \}_{closed}
\end{align*}
Following we show a table type that Typed Lua could assign to a
table constructor that does not include the field $``middlename"$,
but such table type is not subtype of the above table type:
\begin{align*}
\{ ``firstname":\String, ``lastname":\String \}_{closed}
\end{align*}

We introduced \emph{open} table types to allow the refinement
of table types, and also to allow programmers to omit optional
fields in the initialization of table types.
To do that, we defined new subtyping rules that allow \emph{open}
table types to omit fields when $\Nil$ is subtype of the of type
of the value of the omitted key.
The following subtyping rules show this behavior:

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{open|closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v\}_{open|closed}} 
\end{array}
\end{array}
\]

However, we still had the that
\begin{align*}
\{ & ``firstname":\String, \\
   & ``middlename":\String,\\
   & ``lastname":\String \}_{open} \not \subtype \\
\{ & ``firstname":\String, \\
   & ``middlename":\String \cup \Nil,\\
   & ``lastname":\String \}_{closed}
\end{align*}

We introduced \emph{unique} table types to solve this problem.
We use \emph{unique} table types only to denote the type of the
table constructor.
The following subtyping rules allow us to use table constructors
that initialize optional fields:

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE4}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype_{n} v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{unique|open|closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE5}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype_{n} v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v\}_{unique|open|closed}} 
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD4}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype_{n} k_{2}:t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD5}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \Const \; k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD6}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
\end{array}
\end{array}
\]

Notice that we introduced a special subtyping rule to check the
subtyping of \emph{unique} table fields.
The difference is that \emph{unique} table types have width and
depth subtyping on all fields.

Even though \emph{unique} table types solve the problem of using
table constructors to initialize tables that contain optional fields,
we still have the problem when we try to use a local variable that
holds an \emph{open} table to initialize another table that contains
optional fields.

\emph{/* Do we really need closed, open, and unique? */}

\emph{/* Also, can we make these subtyping rules for tables more clear? */}

Table types can be ambiguous.
For instance, $\{1:\Number, \Number:\String\}$ is an ambiguous
table type because the type of the value stored by key $1$ can be
$\Number$ or $\String$, as $1 \subtype 1$ and
$1 \subtype \Number$.
Table types can get even more ambiguous when we use the dynamic
type in the type of table keys.
For instance, the table type $\{1:\Number, \Number:\String, \Any:\Boolean\}$
is ambiguous because the type of the value stored by key $1$ can be
$\Number$, or $\String$, or $\Boolean$, as $1 \subtype 1$,
$1 \subtype \Number$, and $1 \lesssim \Any$.
The type of the value stored by a key of type $\Number$,
which is not the literal type $1$, can be $\Number$ or $\Boolean$,
as $\Number \subtype \Number$, and $\Number \lesssim \Any$.

>>>>

In this chapter we present the formalization of Typed Lua.
Besides its practical contribution, Typed Lua also has some interesting
contributions on the field of optional type systems for scripting
languages.
Namely, it is a simple type system that covers several Lua idioms
and features, such as the refinement of tables, multiple return values,
and optional parameters.
We will use typing rules to explain how we implemented the features of Typed Lua.

@@

Our type system uses subtyping \citep{cardelli1984smi,abadi1996to} to order
types and consistent-subtyping \citep{siek2007objects,siek2013mutable}
to allow the interaction between statically and dynamically typed code.
We will explain the subtyping and consistent-subtyping rules along this
section while we introduce each type from our type system.
We will focus the discussion on the definition of subtyping because,
as we mentioned in Chapter \ref{chap:review}, we can combine the
consistency and subtyping relations to achieve consistent-subtyping.
The differences between subtyping and consistent-subtyping are the way
they handle the dynamic type, and the fact that subtyping is transitive,
but consistent-subtyping is not.

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-level types}}\\
t ::= & \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; p \rightarrow r & \textit{function types}\\
& | \; \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; \pi_{i} & \textit{projection types}\\
l ::= & \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
b ::= & \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
k ::= & \;\; l \; | \; b \; | \; \Any & \textit{key types}\\
v ::= & \;\; t \; | \; \Const \; t & \textit{field types}\\ 
\multicolumn{3}{c}{}\\
\multicolumn{3}{c}{\textbf{Second-level types}}\\
s ::= & \;\; t &\\
& | \; t* & \textit{variadic types}\\
& | \; t \times s & \textit{tuple types}\\
p ::= & \;\; \Void \; | \; s & \textit{parameters list type}\\
r ::= & \;\; \Void \; | \; s \; | \; s \sqcup r & \textit{return types}\\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

Literal types represent the type of literal values.
They can be the boolean values $\False$ and $\True$,
any integer value, any floating point value, and any string value.
We shall see that literal types are important to represet table types
as records.

Our type system also includes the following base types: $\Boolean$,
$\Integer$, $\Number$, and $\String$.
The base types $\Boolean$ and $\String$ represent the values that
during run-time Lua tags as \texttt{boolean} and \texttt{string},
respectively;
the base types $\Integer$ and $\Number$ represent the values
that during run-time Lua tags as \texttt{number}.
We introduced two different base types to represent Lua numbers
because Lua introduced integers in version 5.3.

The subtyping rules for literal types include a reflexive rule,
and the rules for defining that literal types are subtypes of
their respective base types;
the subtyping rules for base types include a reflexive rule,
and another rule for defining that $\Integer$ is subtype of
$\Number$.
We defined these subtyping rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-LITERAL}\\
\senv \vdash l \subtype l
\end{array}
\;
\begin{array}{c}
\mylabel{S-FALSE}\\
\senv \vdash \False \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-TRUE}\\
\senv \vdash \True \subtype \Boolean
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-INT1}\\
\senv \vdash {\it int} \subtype \Integer
\end{array}
\;
\begin{array}{c}
\mylabel{S-INT2}\\
\senv \vdash {\it int} \subtype \Number
\end{array}
\;
\begin{array}{c}
\mylabel{S-FLOAT}\\
\senv \vdash {\it float} \subtype \Number
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-STRING}\\
\senv \vdash {\it string} \subtype \String
\end{array}
\begin{array}{c}
\mylabel{S-BASE}\\
\senv \vdash b \subtype b
\end{array}
\;
\begin{array}{c}
\mylabel{S-NUMBER}\\
\senv \vdash \Integer \subtype \Number
\end{array}
\end{array}
\]

@@

We could have omitted the rules \textsc{S-LITERAL}, \textsc{S-BASE},
and \textsc{S-INT2}, because the first two are reflexive and we
could achieve the third one trhough transitivity.
More precisely, we could have defined subsumption rules that
specify the reflexive and trasitive behavior of subtyping,
and use these rules instead of the ones we defined.
For instance, we would not need \textsc{S-INT2} because we could
use the transisitive subsumption rule to derive that
\[
\dfrac{\senv \vdash 1 \subtype \Integer \;\;\;
       \senv \vdash \Integer \subtype \Number}
      {\senv \vdash 1 \subtype \Number}
\]

@@

The type $\Value$ is the top type, so any first-level type is
subtype of $\Value$.
We shall see that this type along with variadic types help to
drop extra values on multiple assignments, and function calls,
thus preserving the semantics of Lua in these cases.
Following we show its subtyping rule:
\[
\begin{array}{c}
\mylabel{S-VALUE}\\
\senv \vdash t \subtype \Value
\end{array}
\]

The dynamic type $\Any$ is the type that allows programmers to mix
static and dynamic typing.
The type $\Any$ is neither the bottom nor the top type, but a
separated typed that is subtype only of itself:
\[
\begin{array}{c}
\mylabel{S-ANY}\\
\senv \vdash \Any \subtype \Any
\end{array}
\]

@@

The rules that handle the dynamic type are the only rules that
differ between subtyping and consistent-subtyping.
If we set the type $\Any$ as both bottom and top types of our
subtyping relation, then all programs would type check without errors.
This would happen due to the transitivity of subtyping, that is,
we would be able to down-cast any type to $\Any$ and then up-cast
$\Any$ to any type.
While the rule \textsc{S-ANY} is neutral and makes transitivity
not be a problem, the rules \textsc{C-ANY1} and \textsc{C-ANY2}
are the rules that allow the dynamic type to interact with other
first-level types, and thus allow dynamically typed code coexist
with statically typed code.
Because of these two rules consistent-subtyping cannot be transitive.

@@

Union types are disjoint, that is, the types in the union are not
consistent-subtype of any other type in the same union.
For instance, the union type $\Boolean \;\cup\; \Any$ results the
type $\Any$, because $\Boolean \lesssim \Any$;
and the union type $\Number \;\cup\; \Nil \;\cup\; 1$ results the union
type $\Number \;\cup\; \Nil$ because $1 \lesssim \Number$.

\emph{/* Do we need to show how Typed Lua simplifies union types? */}

To introduce function types we first need to introduce
second-level types, as they appear on both sides of function types.
Second-level types are tuples of first-level types
that can optionally end with a variadic type.
Our type system needs second-level types because tuples are
not first-class values in Lua, though they may appear on
argument passing, multiple returns, and multiple assignments.
A variadic type $t*$ is a generator for a sequence of values of
type $t \cup \Nil$, and it represents the type of the vararg
expression.
Next we show their subtyping rules:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VARARG}\\
\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
      {\senv \vdash t_{1}* \subtype t_{2}*}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TUPLE1}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2} \;\;\;
       \senv \vdash s_{1} \subtype s_{2}}
      {\senv \vdash t_{1} \times s_{1} \subtype t_{2} \times s_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TUPLE2}\\
\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \;\;\;
       \senv \vdash t_{1}* \subtype s_{2}}
      {\senv \vdash t_{1}* \subtype t_{2} \times s_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TUPLE3}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil \;\;\;
       \senv \vdash s_{1} \subtype t_{2}*}
      {\senv \vdash t_{1} \times s_{1} \subtype t_{2}*}
\end{array}
\end{array}
\]

The subtyping rules for variadic and tuple types show that
these types are covariant.
The subtyping rule for variadic types is straightforward:
it shows that a variadic type $t_{1}*$ is subtype of another
variadic type $t_{2}*$ if $t_{1} \cup \Nil$ is subtype of
$t_{2} \cup \Nil$.
The first rule for tuple types shows that a tuple type
$t_{1} \times s_{1}$ is subtype of another tuple type
$t_{2} \times s_{2}$ if $t_{1}$ is subtype of $t_{2}$
and $s_{1}$ is subtype of $s_{2}$.
A tuple type can optionally end with a variadic type, and
the rules \textsc{S-TUPLE2} and \textsc{S-TUPLE3} handle this case.
The rule \textsc{S-TUPLE2} shows that a variadic type is subtype of
a tuple type whenever the type of the value that it generates is
subtype of all the elements of the tuple.
The rule \textsc{S-TUPLE3} shows that a tuple type is subtype of a
variadic type whenever all the elements of the tuple are subtype of
the type of the value that it generates.
We shall see that these subtyping relations along with the
types $\Value$ and $\Nil$ allow our type system to reflect
the semantics of Lua on discarding extra parameters and
replacing missing parameters.

In our type system function types have the form $p \rightarrow r$,
where $p$ is the type of the parameters list and $r$ is the
return type.
The type of the parameters list can be $\Void$ or a tuple type,
while the return type can be $\Void$, a tuple type, or a union of
tuple types.
The type $\Void$ means an empty tuple.
Our type system includes union of tuples on the return type because
in Lua it is common to overload the return type to denote error,
as we mentioned in Section \ref{sec:statistics}.

@@

Our type system includes table types to represent Lua tables.
The syntactical form of table types is $\{ k_{i}{:}v_{i}, ..., k_{n}{:}v_{n} \}_{tag}$.
When $n = 0$, then $i = 0$ and it represents the type of an empty table.
When $n > 0$, then $i$ varies from 1 to $n$, each $k_{i}$ represents
the type of a table key, and each $v_{i}$ represents the type of the
value that the table key $k_{i}$ maps to.
We restricted the type of keys to literal types, base types, and the dynamic type,
but we can use any first-level type to represent the type of values they map to.
We made this restriction to the type of the keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The dynamic type is an option when we need a more loosen table type.

@@

The following subtyping rule shows that \emph{closed} table types
have width subtyping, and depth subtyping on $\Const$ fields;
it resembles the subtyping rule for record types.

\[
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
       \senv \vdash v_{j} \subtype_{c} v_{i}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
\end{array}
\]

The rule \textsc{S-TABLE1} introduced the relation $\subtype_{c}$.
It is just an auxiliary relation that defines the subtyping of the
type of the values stored in table fields.
We needed an auxiliary relation because the subtyping of the type
of the values stored in the table fields change according to the
tag of the table type.
For closed tables, subtyping handles mutable fields invariantly,
and imutable fields covariantly.
The following rules define this behavior.
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2} \;\;\;
       \senv \vdash v_{2} \subtype v_{1}}
      {\senv \vdash v_{1} \subtype_{c} v_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\end{array}
\]

There are two limitations on \emph{closed} table types that led us to
introduce \emph{open} and \emph{unique} table types.
First, they do not allow programmers to use the table constructor to
initialize table types that include optional fields.
For instance,
\begin{align*}
\{ & ``firstname":\String, ``lastname":\String \}_{closed} \not \subtype\\
\{ & ``firstname":\String,\\
  &  ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
\end{align*}
Second, they do not allow programmers to use the table construtor to
initialize table types that represent hashes.
For instance,
\begin{align*}
\{ & ``x":\Number, ``y":\Number, ``z":\Number \}_{closed} \not \subtype\\
\{ & \String:\Number \cup \Nil\}_{closed}
\end{align*}

We introduced \emph{open} table types to allow the refinement
of table types, and also to allow programmers to omit optional
fields in the initialization of table types.
To do that, we defined a subtyping rule that allow \emph{open}
table types to omit fields when $\Nil$ is subtype of the type of
the value of the ommited key.
The following subtyping rule show this behavior:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
\end{array}
\]

The rule \textsc{S-TABLE2} introduced the relation $\subtype_{o}$.
It is an auxiliary relation that defines the subtyping relation
on optional fields, and we defined it as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD4}\\
\dfrac{\senv \vdash \Nil \subtype v}
      {\senv \vdash \Nil \subtype_{o} v}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD5}\\
\dfrac{\senv \vdash \Nil \subtype v}
      {\senv \vdash \Nil \subtype_{o} \Const \; v}
\end{array}
\end{array}
\]

Although \emph{open} table types let programmers omit optional fields,
they do not let programmers initialize optional fields that are
mutable, unless they use the same type of the field in the initialization.
For instance,
\begin{align*}
\{ & ``firstname":\String,\\
   & ``middlename":\String, ``lastname":\String \}_{open} \not \subtype \\
\{ & ``firstname":\String, \\
   & ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
\end{align*}

We introduced \emph{unique} table types to solve this problem.
We use \emph{unique} table types to denote the type of the
table constructor.
The following subtyping rule allow programmers to use table constructors
to initialize and omit optional fields:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
\end{array}
\]

The rule \textsc{S-TABLE3} introduced the relation $\subtype_{u}$.
It is an auxiliary relation that makes \emph{unique} table types
have width and depth subtyping on all fields, and we defined it
as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD6}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{u} v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD7}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash \Const \; v_{1} \subtype_{u} \Const \; v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD8}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{u} \Const \; v_{2}}
\end{array}
\end{array}
\]
